Pipit: Reactive Systems in \fstar ~(Extended Abstract)

### 2023-05-16 2023pipit/Main.tex
shuffling

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on a subset of Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.

@@

% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
The implementation starts by looking at the list of coefficients, and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before finally summing the two parts.
At the start of execution the delay is initialised to zero.

Although this is a simple definition, verifying it \emph{and} executing it in Lustre is somewhat impractical, as tool support varies widely.
To write this program in Lustre, we would likely need to reformulate the program to use arrays.
Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays without compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
%TODO defer explanation to related work

Instead, inspired by Bedrock~\cite{chlipala2013bedrock}, we define Pipit as a small core language embedded inside a theorem prover (\fstar{} in our case).
By implementing Pipit as an embedded language, we can express the above FIR filter, but the lists are evaluated away by the meta-language to a simple core program without lists or arrays.
For example, we can compute the weighted moving average with the following concrete coefficients:

@@

Normalising the above definition of $\mbox{fir}_2$ removes all lists.
The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:

@@

\subsection{Bounded input, bounded output}
\label{ss:tut:bibo}
We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range.
Properties like this can occur as an internal invariant in a larger system:

@@

If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is no longer sufficient; we need a stronger invariant.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:
% TODO: intuition, (0 fby x) < n = true fby (x < n)

@@

To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.
The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\error{\textit{error}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}

% Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the examples in \autoref{ss:tut:bibo}, the two properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example.

\section{Related work}

>>>>

To be confident that our proofs hold on the real system, we need a single semantics that is shared between the compiler and the model-checker or prover.
In this abstract we introduce Pipit\footnote{Implementation available at \url{https://github.com/songlarknet/pipit}}, an embedded domain-specific language for implementing and verifying controllers in \fstar{}.
Pipit aims to provide a high-level language based on Lustre, while reusing \fstar{}'s proof automation and manual proofs for verifying controllers~\cite{martinez2019meta}, and using \lowstar{}'s C-code generation for real-time execution~\cite{protzenko2017verified}.
% By designing Pipit as an embedded language, we can reuse many of the \fstar{} meta-language's features while retaining a small core language.
Pipit translates its expression language to a transition system for k-inductive proofs, which is verified; verifying the translation to imperative code is ongoing.

@@

% Our FIR filter takes two arguments, one describing the coefficients and one describing the signal to filter.
The coefficient vector is represented by a list of reals, while the signal is a \emph{stream} of reals, and the result is the filtered stream.
% \footnote{The actual implementation uses integers with fixed-point arithmetic, but we use reals here to reduce clutter.}
The implementation starts by looking at the list of coefficients and returns zero if the list is empty.
If the list is not empty, then we multiply the most recent value of the stream by the coefficient ($\signal \cdot c$); we also take the result of applying the remaining coefficients to the signal stream ($\mbox{fir~} \cs ~ \signal$) and delay it ($\xfby{0}{\ldots}$) before summing the two parts.
At the start of execution the delay is initially zero.

Pipit is an \emph{embedded} language, like Bedrock~\cite{chlipala2013bedrock}: in this example, the stream type denotes an actual Pipit expression, while the list type and its associated pattern match is part of the \fstar{} meta-language.
To get the real Pipit program, we need to apply the filter to a concrete list of coefficients:

@@

Normalising this definition evaluates away all of the lists.
The result fits in the core language defined in \autoref{s:core}, for which we can generate real-time imperative code:

@@

\subsection{Bounded input, bounded output}
\label{ss:tut:bibo}
We can now state a \emph{bounded-input-bounded-output} (BIBO) property, which says that if the inputs have always been within some particular range, then the outputs are also within the range:
% Properties like this can occur as an internal invariant in a larger system:

@@

If we wish to prove a similar BIBO property for a filter with more coefficients, standard induction over the transition system is not sufficient; the relationship between the stacked delays in the filter and sofar is not clear from a single step of the transition system.
One simple automated way to strengthen invariants is via \emph{k-induction}~\cite{hagen2008scaling}, which adds more context by assuming that the property holds for $k$ previous steps of the transition relation.
We can define analogous functions $\fir_3$ and $\bibo_3$ which operate on the coefficients $[0.7; 0.2; 0.1]$, and use k-induction for $k = 2$ as follows:

@@

To demonstrate the feasibility of Pipit, we have implemented and verified a simple controller.
This system controls a water-flow solenoid to fill the reservoir of a coffee machine and includes multiple safeguards to reduce the risk of flooding.
The controller has two boolean inputs: the \emph{stop} switch and the \emph{low level} indicator; it returns a boolean indicating whether to engage the solenoid.
The stop switch indicates whether the reservoir's lid is open or closed; the system should never operate while the lid is open as water could spill out.
The controller should not allow water to flow for more than a minute as this may indicate a leak; if so, the controller enters a terminal error state.
Finally, to avoid switching the solenoid too often, the controller waits for ten seconds of low water level before trying to engage.

\newcommand\estop{\textit{stop}}
\newcommand\low{\textit{low}}
\newcommand\soltry{\textit{try}}
\newcommand\error{\textit{error}}
\newcommand\solen{\textit{engage}}
\begin{tabbing}
  @MM@\= @let@ engage \= @MMMM@ \= \kill
  @let@ $\mbox{reservoir}$ (\estop{} \low{}: stream $\BB$): stream $\BB$ = \\
  \> @let@ \soltry{} \> = true\_for \textit{TEN\_SECONDS} (not \estop{} $\wedge$ \low{}) @in@ \\
  \> @let@ \error{} \> = any (true\_for \textit{ONE\_MINUTE} \soltry{}) @in@ \\
  \> @let@ \solen{} \> = \soltry{} $\wedge$ not \error{} @in@ \\
  \> @check@ (\solen{} $\implies$ not \estop{}); \\
  \> @check@ (\solen{} $\implies$ \low{}); \\
  \> \solen{}
\end{tabbing}
\pagebreak

% Here, \emph{true\_for t} is time-bounded past-globally and \emph{any} is past-finally.
As with the examples in \autoref{ss:tut:bibo}, the two properties can be automatically verified.
% by \fstar{}'s proof automation.
Pipit generates real-time C code for this example\footnote{For a video of the controller in action, see \url{https://youtu.be/6IybbQFPOl8}}.

\section{Related work}
\label{s:related-work}
% TODO arrays

Although the FIR filter from \autoref{s:tut} is a simple definition, verifying it \emph{and} executing it with existing Lustre tools is impractical.
To write this program in Lustre we would likely need to reformulate the program to use arrays, as dynamically-allocated data structures are not well-suited to real-time execution.
Vélus does not support arrays~\cite{bourke2017formally}; Kind2 uses a custom syntax for arrays with no compiler support~\cite{champion2016kind2}; and Lustre V6 does support arrays~\cite{jahier2016lustre}, but its model-checker Lesar cannot reason about integers or reals~\cite{raymond2008synchronous}.
